---
title: "Markdown"
author: "Alireza Moradi"
date: "8/16/2019"
output: html_document
---
# Analyzing damage on public health and economy 
#### Caused by severe weather events
A brief abalysis on the amount of harm done by different severe weather events on public health and economic consequences.

The basic goal of this paper is to explore the NOAA Storm Database and answer some basic questions about severe weather events.
The below analysis was run on a MacBook Pro, 2.9Ghz intel core i7 cpu, MacOs Mojave 10.14.6.

```{r setup, include=FALSE,echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Processing
Loading required packages and the data into R.
```{r,message=FALSE}
library(R.utils)
library(data.table)
library(dplyr)
library(ggplot2)
```
```{r,cache=TRUE,message=FALSE}
data <- fread("02_course-project-2_repdata-data-StormData.csv.bz2", sep = ",", header = T)
```

# Questions
### 1. Across the United States, which types of events are most harmful with respect to population health?
```{r,cache=TRUE}
fatalities <- data %>% select(EVTYPE,FATALITIES) %>% group_by(EVTYPE) %>% summarise(sum = sum(FATALITIES)) %>% arrange(-sum)
print(fatalities)
```
Total number of fatalities caused by each type of Event.

```{r,cache=TRUE}
injuries <- data %>% select(EVTYPE,INJURIES) %>% group_by(EVTYPE) %>% summarise(sum = sum(INJURIES)) %>% arrange(-sum)
print(injuries)
```
Total number of injuries caused by each type of Event.

### 2. Across the United States, which types of events have the greatest economic consequences?

```{r,cache=TRUE}
w <- function(k){
    for(EXP in k){
        if(toupper(EXP) == "K"){
            k <- as.numeric(10^3)
        }
        else if(toupper(EXP) == "M"){
            k <- as.numeric(10^6)
        }
        else if(toupper(EXP) == "B"){
            k <- as.numeric(10^9)
        }
        else if(toupper(EXP) == "H"){
            k <- as.numeric(10^2)
        }
        else if(EXP == "?" | EXP == "" | EXP == "-"){
            k <- as.numeric(0)
        }
        else{
            k <- as.numeric(0)
        }
    }
    k
}
data$PROPDMGEXP <- sapply(X = data$PROPDMGEXP,FUN = w,simplify = "array")
data$CROPDMGEXP <- sapply(X = data$CROPDMGEXP,FUN = w,simplify = "array")
data$PROPDMG <- data$PROPDMG * data$PROPDMGEXP
data$CROPDMG <- data$CROPDMG * data$CROPDMGEXP

temp <- data %>% group_by(EVTYPE) %>% summarise(sum = sum(PROPDMG,CROPDMG)) %>% arrange(-sum)
print(temp)
```
Total economy impact seperated by Events.

# Results
### Public Health Impact
```{r,fig.align='center'}
g <- ggplot(fatalities[1:10,],aes(reorder(factor(EVTYPE),-sum),sum))
g <- g + geom_bar(stat = "identity") + ylim(c(0,6000))
g <- g + theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1),plot.title = element_text(hjust = 0.5))
g <- g + labs(x = "Event",y = "Total fatalities",title = "Top 10 Events with highest total fatalities")
print(g)
```

According to the above plot, Tornado has the highest total fatalities followed by Excessive Heat and Flash Flood.

```{r,fig.align='center'}
g <- ggplot(injuries[1:10,],aes(reorder(factor(EVTYPE),-sum),sum))
g <- g + geom_bar(stat = "identity") + ylim(c(0,100000))
g <- g + theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1),plot.title = element_text(hjust = 0.5))
g <- g + labs(x = "Event",y = "Total injuries",title = "Top 10 Events with highest total injuries")
print(g)
```

According to the above plot, Tornado caused around 91000 injuries which is the highest, Followed by TSTM Wind and FLood.

#### Conclusion
Tornado has the highest health impact altogether.

### Economy Impact
```{r,fig.align='center'}
g <- ggplot(temp[1:10,],aes(reorder(factor(EVTYPE),-sum),sum))
g <- g + geom_bar(stat = "identity")
g <- g + theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1),plot.title = element_text(hjust = 0.5))
g <- g + labs(x = "Event",y = "Total economy impact($USD)",title = "Top 10 Events with highest economic impact")
print(g)
```

According to the above plot, Flood caused around "150 BILLION $" of damage, Which is the highest economy impact among all others.